FITS format for planetary surfaces: definitions, applications and best practices.

### 2018-02-01 epss_marmo.tex

\section{Introduction}
\label{sec:intro}
Planetary science brings together several research communities
(geologists, astronomers, physicists, geochemists, etc.),
and planetary missions produce a growing amount of heterogeneous data.
To efficiently extract scientific information from those data, they must
be made integrable, thus inter-operable.
Openness of data format and processing techniques is becoming a necessity,
to guarantee the reproducibility of the results obtained from those data.

This paper proposes to use Flexible Image Transport System\footnote{https://fits.gsfc.nasa.gov/}
(FITS) \citep{fitsorig,fitsver3} in planetary surface investigations,
and describes how FITS metadata can easily be inserted in the
PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
metadata distribution model.
FITS format is open and flexible, largely implemented in open and efficient
processing tools.
Those reasons make FITS format appropriate for large amounts of raster data
processing.
The goal of this approach is to make easier data mining and re-processing
in planetary surface investigations, promoting general software based on those
standards.
The option to use FITS within the planetary domain could potentially
homogenize methods from acquisition, to visualization, while
giving chances to optimize data processing.

This paper is organized as follows:
\begin{enumerate}
\item{a brief description of FITS format;}
\item{a proposition for planetary surface metadata description in FITS;}
\item{dictionaries between FITS and PDS metadata description;}
\item{some examples of planetary dataset in FITS;}
\item{current developments for making FITS usable in planetary sciences.}
\end{enumerate}

@@

\subsection{FITS format}
FITS stands for 'Flexible Image Transport System'.
It is an open digital standard, defined for data acquisition and archiving
in astronomical observatories back in the late 70's, and it is used for
spatial telescope data.
In fact, FITS is the IAU-approved standard format for astronomical
data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard implemented formats in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
Finally, FITS format is compatible with the Planetary Data System (PDS)
archiving specifications and supported by a large number of open libraries
and software tools.

FITS is already able to propose standard formatting for some data products quite
common by now in planetary surface investigations.
In particular, Multi-Extension FITS (MEF) schema proposes an easy way
to store inhomogeneous digital information (reflectance, calibration data, vector
table data, etc.) in the same file each with relative metadata, as well as
multi-detector imagery (e.g. from HiRISE \citep{hirise}) or hyperspectral cubes
(e.g. from CRISM \citep{crism} or OMEGA \citep{omega}).
FITS has been already chosen to distribute, e.g.,
Hayabusa\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
Akatsuki\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
and some of the Dawn\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
To be efficiently used in planetary surface investigations, FITS metadata must
be extended in order to take into account the size of the reference body.
Planetary body shapes and orientations are well defined and standardized by the
International Astronomical Union (IAU) Working Group on Cartographic
Coordinates and Rotational Elements (WGCCRE).
This group reports triennially on the preferred rotation rate, spin axis, prime meridian,
and reference surface for planets and satellites which helps ensure that
cartographic endeavors are effectively comparable \citep{wgccre}.
In the framework of the VESPA\footnote{http://europlanet-vespa.eu/} \citep{vespapss}
component of the EuroplanetRI2020 project an extension to FITS meta-
data (GeoFITS) has been proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention}.

\section{Planetary Surface proposed convention}
When willing to use FITS format for Planetary Surfaces the issue of rigorous
geographical representation must be addressed.
Planetary surface researchers for their mapping and analysis needs, commonly use
Geographical Information Systems (GIS).
In general, GIS applications excel in data interoperability even
though some have historically been anchored to Earth's spatial description.
FITS World Coordinate System representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html.}
simplifies the spatial coordinate description with respect to historical terrestrial
references.
There is no need for oblique projection definitions \citep{Snyder1987,calagreisenI}
as projection parameters (projection center or reference point)
describe the difference between the Mercator and
Oblique Mercator or Simple Cylindrical and
Equirectangular map projections.
There is no need for East or West positive longitude definitions as the pixel
conversion in world coordinates is set by a common
oriented matrix.

@@

\begin{table}
\label{table:proj}
\caption{Projection names and codes already available in FITS suitable for planetary applications:
their translation in PDS vocabulary is also provided}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Sanson-Flamsteed & SFL & Sinusoidal \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Plate Carre & CAR & Equirectangular \\
Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
Mercator & MER & Mercator, Oblique Mercator \\
Ortographic & SIN & Orthographic \\
Zenithal Perspective & AZP & General Vertical Near-sided Perspective \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}}
\end{tabular}
\end{table}

>>>>

\section{Introduction}
\label{sec:intro}
Planetary science brings together several research communities
(geologists, astronomers, physicists, geochemists, etc.),
and planetary missions produce a growing amount of heterogeneous data.
To efficiently extract scientific information from those data, they must
be made integrable, thus interoperable.
Openness of data format and processing techniques is becoming a necessity,
to guarantee the reproducibility of the results obtained from those data.

This paper proposes to use Flexible Image Transport
System\footnote{https://fits.gsfc.nasa.gov/}
(FITS) format \citep{fitsorig,fitsver3} in planetary surface investigations,
and describes how FITS metadata can easily be inserted in the
PDS4\footnote{https://pds.nasa.gov/tools/standards-reference.shtml}
metadata distribution model.
FITS format is open and flexible, largely implemented in open and efficient
processing tools.
Those reasons make FITS format appropriate for large amounts of raster data
processing.
The goal of this approach is to make easier data mining and reprocessing
in planetary surface investigations, promoting general software based on those
standards.
The option to use FITS within the planetary domain could potentially
homogenize methods from acquisition, to visualization, while
giving chances to optimize data processing.

This paper is organized as follows:
\begin{enumerate}
\item{a brief description of FITS format;}
\item{a proposition for planetary surface meta-data description in FITS;}
\item{dictionaries between FITS and PDS meta-data description;}
\item{some examples of planetary dataset in FITS;}
\item{current developments for making FITS usable in planetary sciences.}
\end{enumerate}

@@

\subsection{FITS format}
FITS is an open digital standard, defined for data acquisition and archiving
in astronomical observatories in the late 70's, and it is used for
spatial telescope data.
The International Astronomical Union approved FITS as the standard format
for astronomical data\footnote{https://fits.gsfc.nasa.gov/iaufwg/history/IAU\_1982\_resolution\_c1.html}. 
Therefore, FITS is one of the standard implemented formats in the Virtual
Observatory\footnote{http://www.ivoa.net/astronomers/using\_the\_vo.html}
(VO).
Finally, FITS format is compatible with the Planetary Data System (PDS)
archiving specifications and supported by a large number of open libraries
and software tools.

FITS is already able to propose standard formatting for some data products
quite common by now in planetary surface investigations.
In particular, the Multi-Extension FITS (MEF) schema proposes an easy way
to store inhomogeneous digital information (reflectance, calibration data,
vector, table data, etc.) in the same file, each with relative metadata,
as well as multi-detector imagery (e.g. from HiRISE \citep{hirise})
or hyperspectral cubes with geometry information (e.g. from CRISM
\citep{crism} or OMEGA \citep{omega} instruments).
FITS has been already chosen to distribute, e.g.,
Hayabusa\footnote{http://darts.isas.jaxa.jp/planet/project/hayabusa/index.html},
Akatsuki\footnote{http://darts.isas.jaxa.jp/planet/project/akatsuki/index.html}
and some of the Dawn\footnote{https://sbn.psi.edu/pds/archive/dawn.html} data.
To be efficiently used in planetary surface investigations, FITS metadata must
be extended in order to take into account the size of the reference body.
Planetary body shapes and orientations are well defined and standardized by
the IAU Working Group on Cartographic Coordinates and Rotational Elements
(WGCCRE).
This group reports triennially \citep{wgccre} on the preferred rotation rate,
spin axis, prime meridian, and reference surface for planets and satellites
ensuring that cartographic endeavors are effectively comparable.
In the framework of the VESPA\footnote{http://europlanet-vespa.eu/}
\citep{vespapss} workpackage of the EuroplanetRI2020\footnote{http://www.europlanet-2020-ri.eu/}
project an extension to FITS metadata (GeoFITS) has been
proposed\footnote{https://voparis-confluence.obspm.fr/display/VES/GeoFITS:+Planetary+Data+FITS+format+and+metadata+convention},
taking into account the WGCCRE recommendations.

\subsection{Planetary Surface proposed convention}
\label{ssec:geofits}
When willing to use FITS format for planetary surfaces the issue of rigorous
geographical representation must be addressed.
For their mapping and analysis needs planetary researchers commonly use
Geographical Information Systems (GIS).
In general, GIS applications excel in data interoperability even
though some have historically been anchored to Earth's spatial description.
FITS World Coordinate System
representation\footnote{http://fits.gsfc.nasa.gov/fits\_wcs.html}
simplifies the spatial coordinate description with respect to historical
terrestrial references.
There is no need for oblique projection definitions \citep{calagreisenI,Snyder1987}
as projection parameters (projection center or reference point)
describe the difference between,e.g., the Mercator and
Oblique Mercator or Simple Cylindrical and
Equirectangular map projections.
There is no need for East or West positive longitude definitions as the pixel
conversion in world coordinates is set by an oriented matrix.

\begin{table}
\label{table:proj}
\caption{Projection names and codes already available in FITS suitable for
planetary applications: their translation in PDS4 vocabulary is
provided. See Table A.1. in \citet{calagreisenI} for details.}
\centering
\begin{tabular}{l l l}
\hline
FITS Name$^{a}$ & FITS code$^{a}$ & PDS name$^{b}$ \\ 
\hline
Sanson-Flamsteed & SFL & Sinusoidal \\
Zenithal Equal-Area & ZEA & Lambert Azimuthal Equal Area \\
Conic Ortomorphic & COO & Lambert Conformal Conic\\
Plate Carre & CAR & Equirectangular \\
Cylindrical Equal-Area & CEA & \textit{there is no CEA option in PDS4 neither in ISIS3}\\
Mercator & MER & Mercator, Oblique Mercator \\
Ortographic & SIN & Orthographic \\
Zenithal Perspective & AZP & General Vertical Near-sided Perspective \\
Stereographic & STG & Stereographic, Polar Stereographic \\
Gnomonic & TAN & Gnomonic \\
\hline
\multicolumn{3}{l}{$^{a}$From \citet{calagreisenI}} \\
\multicolumn{3}{l}{$^{b}$From the PDS4 Data Dictionary\footnote{https://pds.jpl.nasa.gov/pds4/doc/dd/current/PDS4\_PDS\_DD\_1900.pdf}}
\end{tabular}
\end{table}

@@
